---
title: "Data Exploration"
output: html_document
---

## Working with data frames

The following code chunks contain examples for importing and visualizing data.

In the code chunk below, I first loaded my packages:

1. `tidyverse`
2. `here`

Next, I imported my data set called `clams`. I then sourced `my_function.R` from my R directory. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)

clams <- read.table(here::here("data","Clams_Zuur-et-al-2009.txt"),
                    stringsAsFactors = FALSE, header = TRUE)

source(here::here("R","my_function.R"))

file.exists(here::here("data","Clams_Zuur-et-al-2009.txt"))
```

```{r data_create}
example_data <- data.frame(x = 1:20,
                           y = rnorm(20),
                           group = rep(c("a","b"), each = 10))
```

```{r data_plot}

plot(x = example_data$x, y = example_data$y)
plot(y ~ x, data = example_data)

```

```{r data_ggplot}
ggplot(data = example_data, aes(x = x, y = y, color = group)) +
  geom_point() +
  geom_line()

ggplot(data = example_data, aes(x = x, y = y, group = group)) +
  geom_point() +
  geom_line()
```

```{r ggboxplot}
ggplot(data = example_data, aes(x = group, y = y)) +
  geom_boxplot() +
  geom_point()
```

## Wedge Clam Data

Zuur adds columns using the `$` syntax. I also created the same data.frame (named `log_clams`) using the `mutate()` function from `dplyr`.

```{r}

clams$LNAFD <- log(clams$AFD)
clams$LNLENGTH <- log(clams$LENGTH)


log_clams <- clams %>% 
  mutate(LNAFD = log(AFD),
         LNLENGTH = log(LENGTH))
```


