---
title: "Tidy modeling with broom and purrr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyverse)
library(nlme)
library(broom.mixed)
```

### Read in data

```{r}

# Read in data
RIKZ <- readr::read_delim(here::here("data/RIKZ.txt"), delim = "\t") %>% 
  
#Convert character vector (Beach) to factor (fBeach) as new column
  mutate(fBeach = factor(Beach)) 
```

#### Linear regression with RIKZ

First let's fit a linear regression model relating `NAP` and `Richness` using `lm()`.

```{r}
lr_mod <- 
  lm(Richness ~ NAP, data = RIKZ)
```

Using the `augment` function from the package `broom`, we can easily return model predictions in a format that is compliant with other tidyverse packages like `ggplot2` and `dplyr`. Note the use of the pipe operator ` %>% ` to pass the model object to the `augment` function.

```{r}
fitted_lr_mod <- lr_mod %>% augment()

#View model fit
str(fitted_lr_mod)
```

We can now pass our fitted model output to `ggplot2` for visualization.

```{r}
lr_mod_plot <- 
  ggplot(data = fitted_lr_mod) +
  geom_line(aes(x = NAP, y = .fitted)) +
  geom_point(aes(x = NAP, y = Richness))

lr_mod_plot
```

Now visualize with confidence intervals.
```{r}
lr_mod_plot +
  geom_ribbon(aes(x = NAP,
                  ymin = .fitted - 2 * .se.fit,
                  ymax  = .fitted + 2 * .se.fit),
              alpha = 0.2)
```


Q1: Combine the original `RIKZ` data set with our fitted output in `ggplot2` to color points by sampling site (i.e. `fBeach`)
```{r}

```

#### Mixed effects models

Fit a random intercept model and return fitted values using `broom.mixed` package.

```{r}
Mlme1 <- 
  lme(Richness ~ NAP, random = ~1|fBeach, data = RIKZ)
```

View the model summary in `tidy` format (i.e. a `data.frame`) using the `broom` functions `tidy()` and `glance()`. The 

```{r}
broom.mixed::tidy(Mlme1)
broom.mixed::glance(Mlme1)
```

The package `broom.mixed` allows us to extend our `augment` function to deal with mixed-effects models. The usage is the same as the linear regression case.

```{r}
Mlme1_fitted <- broom.mixed::augment(Mlme1)
```

Plot the data using `ggplot2`. 

```{r}
ggplot(data = Mlme1_fitted) +
  geom_line(aes(x = NAP, y = .fitted, group = fBeach, color = fBeach)) +
  geom_point(aes(x = NAP, y = Richness, group = fBeach, color = fBeach))
```

Q2: Use `broom` and `ggplot2` functions to replicate and visualize the random intercept and slope model first given on pg. 109 of Zuur et al. 2009. 

```{r}

```

Q3: Repeat the above exercise but for the random effects model specified on pg. 111.

```{r}

```

#### Tidy vectorization of model fitting using purrr**

```{r}

#linear regression example

# Read in some data and clean it up with dplyr
teethN <- readr::read_delim(here::here("data/TeethNitrogen.txt"), delim = " ") %>% 

  #Data were read in with spaces in strings (for example " M2679/93"). We use stringr::str_remove to remove
  #empty values ('\\s+' is a regular expression representing "one or more spaces"). We convert to numeric
  #if necessary.
  mutate(`15N` = as.numeric(str_remove(`15N`, "\\s+")),
         Age = as.numeric(str_remove(Age, "\\s+")),
         Tooth = str_remove(Tooth, "\\s+"))


teethN_fitted <- 
  teethN %>% 
  split(.$Tooth) %>% #Split out 15N values by "Tooth"
  map(~ lm(`15N` ~ Age, data = .)) %>%  #Fit a linear model to each group
  map(augment) #Use augment to return model fits


# purrr::map returns lists. Here we loop through the lists to add in identifying columns. Anyone have a better method? 
for (i in names(teethN_fitted)){
  teethN_fitted[[i]] <- teethN_fitted[[i]] %>% mutate(Tooth = i)
}

#Combine lists into a data.frame for passing to ggplot2
teethN_fitted_fin <- teethN_fitted %>% bind_rows()

#Visualize output
ggplot(data = teethN_fitted_fin) +
  geom_line(aes(x = Age, y = .fitted)) +
  geom_point(aes(x = Age, y = X15N), size = 0.1) +
  facet_wrap(.~Tooth) +
  ylab("15N") +
  ggtitle("Tidy modeling with purrr and broom")
```

** Note: I wouldn't consider this code entirely "tidy" because I used a for loop to add identifying columns to the lists of fitted model outputs. There's no doubt a better way.

Q4: Use the above framework to model and visualize the `TeethNitrogen` data set using GAMs.

```{r}

```

