---
title: ""
runtime: shiny
output: html_document
font: Helvetica
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyverse)
library(nlme)
library(broom.mixed)
library(ggplot2)

RIKZ <- readr::read_delim(here::here("data/RIKZ.txt"), delim = "\t") %>% 
  mutate(fBeach = factor(Beach))
RIKZ$fBeach <- factor(RIKZ$Beach)


RandomInterceptModel <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModel.jpg")
RandomInterceptModelAssumptions <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModelAssumptions.jpg")
RandomInterceptModelVar <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModelVar.jpg")
RandomInterceptModelVar2 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModelVar2.jpg")
RandomInterceptModelRes1 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModelRes1.jpg")
RandomInterceptModelRes2 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomInterceptModelRes2.jpg")

```

<center>

## Mixed Effects Modelling {.tabset .tabset-fade}

<br>

### Random Intercept Model

Suppose we want to ask the following question: <br><b>What is the correlation between two observations from the same beach, and from different beaches?</b>

![](`r RandomInterceptModel`){width=500}

<br><br>
To do this, we simply need to calculate the <b>variance</b> of our model.
<br>
Let's first review the assumptions of our model.
<br>

![](`r RandomInterceptModelAssumptions`){width=350}

<br><br>

Using the independence assumption, we can easily calculate the variance.

![](`r RandomInterceptModelVar`){width=500}

<br><br>

Manually calculating this, we get:

![](`r RandomInterceptModelVar2`){width=500}

<br><br>

This matrix represents our variance and covariance values.

![](`r RandomInterceptModelRes1`){width=500}

<br><br>

And from there we can calculate the <b>correlation</b> value.

![](`r RandomInterceptModelRes2`){width=500}

<br><br>

We can use this value to calculate the <b>design effect</b>, which helps us know whether or not our standard errors are precise (by checking the independence of our results and calculating the <b>effective sample size</b>). Since the variance for the random intercept is <font color="blue">2.94<font color="black">,
and the variance for the residuals is <font color="blue">3.06<font color="black">,
we can use these values to calculate the <b>induced correlation:</b>

2.94<sup>2</sup> / (2.94<sup>2</sup> + 3.06<sup>2</sup>) = 0.48

This is the correlation between observations from the same beach. We can use this value to calculate the design effect (1+(n-1)*p, where p = induced correlation), which helps us check if our standard errors will be skewed.

<br><br><br><br><br><br><br><br><br><br>

```{r echo=FALSE, message=F, warning = F, warnings = F}
RandomSlopeModel <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModel.jpg")
RandomSlopeModelAssumptions <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModelAssumptions.jpg")
RandomSlopeModelVar <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModelVar.jpg")
RandomSlopeModelVar2 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModelVar2.jpg")
RandomSlopeModelVar3 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModelVar3.jpg")
RandomSlopeModelRes1 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeModelVar4.jpg")
RandomSlopeVarCalc1 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeVarCalc1.jpg")
RandomSlopeVarCalc2 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeVarCalc2.jpg")
RandomSlopeVarCalc3 <- here::here("presentations/9-24-Presentation/images",
                                   "RandomSlopeVarCalc3.jpg")
```

### Random Intercept and Slope Model

Suppose we want to ask the following question: <br><b>What is the correlation between two observations from the same beach, and from different beaches?</b>

![](`r RandomSlopeModel`){width=500}

<br><br>
To do this, we need to calculate the <b>variance</b> of our model.
<br>
Let's review the assumptions of our model.
<br>

![](`r RandomSlopeModelAssumptions`){width=350}

<br><br>

Using the independence assumption again, we calculate the variance.

![](`r RandomSlopeVarCalc1`){width=550}
![](`r RandomSlopeVarCalc2`){width=550}

<br><br>
Let's compare our results for variance with the textbook:

![](`r RandomSlopeVarCalc3`){width=550}

<br><br>

Which translates into:

![](`r RandomSlopeModelVar`){width=550}

<br><br>

where, for the following random variables,

![](`r RandomSlopeModelVar2`){width=200}

the following is true:

![](`r RandomSlopeModelVar3`){width=600}

<br><br>

This gives us the following result for our correlation value:

![](`r RandomSlopeModelRes1`){width=600}

<br><br><br><br><br><br><br><br><br><br>


```{r echo=FALSE, message=F, warning = F, warnings = F}
MarginalModel1 <- here::here("presentations/9-24-Presentation/images",
                                   "MarginalModel1.jpg")
MarginalModel2 <- here::here("presentations/9-24-Presentation/images",
                                   "MarginalModel2.jpg")
MarginalModelCov <- here::here("presentations/9-24-Presentation/images",
                                   "MarginalModelCov.jpg")

```

### The Marginal Model

Let's revisit our method for finding the covariance matrix and see if we can get the same results without having to derive the complicated formulas from before.

<br>

One way we can do this is by assuming <b>no random effects</b> and that <b>the residuals of the samples from the same beach are not independent</b>:

![](`r MarginalModel1`){width=500}

<br><br>

In removing the random effects, our new model becomes the regular linear regression model:

![](`r MarginalModel2`){width=500}

<br><br>

From here we can then introduce the kind of depedence structure we would like:

![](`r MarginalModelCov`){width=500}

<br><br>

The following R code is used to find the value of phi:

```{r, echo=T}
M.gls <- gls(Richness ~ NAP, method = "REML", correlation = corCompSymm(form = ~1 | fBeach),
             data = RIKZ)

summary(M.gls$modelStruct)

# Residual standard error
M.gls$residuals[1]
```

<br>

Because this is the parameter estimate of the <b>correlation structure</b>, we use the formula for correlation above to calculate phi directly by multiplying `0.481` by `4.534`.

<br>

But how have we gone about approximating the values we want?

<br><br><br><br><br><br><br><br><br><br>

```{r echo=FALSE, message=F, warning = F, warnings = F}
MLE1 <- here::here("presentations/9-24-Presentation/images",
                                   "MLE1.jpg")
MLE2 <- here::here("presentations/9-24-Presentation/images",
                                   "MLE2.jpg")
MLE3 <- here::here("presentations/9-24-Presentation/images",
                                   "MLE3.jpg")
MLE4 <- here::here("presentations/9-24-Presentation/images",
                                   "MLE4.jpg")
BlockDiagonalForm <- here::here("presentations/9-24-Presentation/images",
                                   "BlockDiagonalForm.jpg")
REML1 <- here::here("presentations/9-24-Presentation/images",
                                   "REML1.jpg")
REML2 <- here::here("presentations/9-24-Presentation/images",
                                   "REML2.jpg")
```

### ML and REML

We will go through the basic set up for how <b>Maximum Likelihood Estimation</b> is done.<br>
Before we review some assumptions, note the fact that we'll assume each beach has only one sample for notational simplicity. (Later we will specify that all Y values will be incorporated into a single column vector):

![](`r MLE1`){width=650}

<br><br>

Since all of the Yi's are independent, the joint probability is the following:

![](`r MLE2`){width=650}

<br><br>

Because we know the observed values for Yi and Xi, we can consider this function as the <b>likelihood function</b> of the unknown parameters:

![](`r MLE3`){width=650}

<br><br>

If we want to find the value for any of the parameters that maximizes the likelihood function, we do the following, with an example below.

![](`r MLE4`){width=300}

<br><br>

```{r echo=FALSE, message=F, warning = F, warnings = F}
library(colourpicker)
fluidPage(
fluidRow(
    column(12,align="center",numericInput("C", "Beta1", value = -10, min = -10, max = 10))
  )
)
```


```{r echo=FALSE, message=F, warning = F, warnings = F}
fluidPage(
fluidRow(column(6,
renderPlot({
  
  x <- y <- seq(-10, 10, length= 30)
  f <- function(x,y){ z <- - x**2 - y**2 }
  z <- outer(x,y,f)
  plot <- persp(x, y, z, theta = 30, phi = 30, expand = 0.5, col = "lightblue",
                      xlab = "Beta1", ylab = "Beta2", zlab = "L(Yi | Beta1, Beta2)",
                main = "Likelihood Function of Beta1 and Beta2")
  c = input$C
  points (trans3d(x = c, y = 0, z = -c**2, pmat = plot), col = 2, pch = 19)
  lines (trans3d(x = x, y = 0, z = (-2*c)*(x-c) - c**2, pmat = plot), col = 2)
  
  
}, height=400)
),
column(6,
renderPlot({
  
  c <- input$C
  fun.1 <- function(x) -x^2
  fun.2 <- function(x) (-2*c)*(x-c) - c**2
  ggplot(data = data.frame(x = 0), mapping = aes(x = x)) + 
    stat_function(fun = fun.1, color="blue") + xlim(-10,10) + ylim(-100,25) + 
    geom_point(aes(x = c, y = fun.1(c)), color="red") + 
    stat_function(fun = fun.2, color="red") + 
    xlab("Beta1") + ylab("L(Yi | Beta1)") + ggtitle("Partial Derivative of Beta1") +
    theme(plot.title = element_text(hjust = 0.5))


  }, height=400)
),

column(12,
renderText({
  
  c <- input$C
  fun.2 <- function(x) (-2*c)
  
  print(paste("The slope (partial derivative of Beta1) at this point is:", fun.2(c)))

})
)
)
)
```

<br><br>

In words, we treat all other parameters as constants and derive our formula to find the value for our parameter that has the highest likelihood.

<br>

<b>However</b>, a problem with this approach is that we are holding constant other parameters (namely, the mean) that are themselves estimates, meaning that our <b>Maximum Likelihood Estimator</b> is a <b>biased estimate</b>. We can reduce this bias by increasing our sample size, and may want to perform a log-likelihood test (which requires ML), but to find a better estimate of the covariance matrix we can use <b>REML</b>. Essentially, we are removing degrees of freedom in proportion to our fixed effects, and apply ML to our regression's least square residuals. 

<br>

Here, we use the stacking method mentioned before that treats all samples in one column, and the errors as a diagonal block matrix, and again adopt the marginal model:

![](`r REML1`){width=550}

<br><br>

More specifically, the matrix A is orthogonal to matrix X, meaning that the X term in the model is completely removed, while A still modifies the covariance matrix V. <br>
The covariance matrix V is of the form:

![](`r BlockDiagonalForm`){width=250}

where Ai corresponds to the covariance matrix of beach i.

<br><br>

Now, once we perform MLE on these random variables Yi, we will get <b>unbiased estimators for the parameters within V</b>. Let's see this in R code.

```{r, echo=T}
RIKZ$fExp <- RIKZ$Exposure
RIKZ$fExp[RIKZ$Exposure == 8] <- 10
RIKZ$fExp <- factor(RIKZ$fExp, levels = c(10,11))
M0.ML <- lme(Richness ~ NAP, data = RIKZ, random = ~1 | fBeach, method = "ML")
M0.REML <- lme(Richness ~ NAP, random = ~1 | fBeach, method = "REML", data = RIKZ)
M1.ML <- lme(Richness ~ NAP + fExp, data = RIKZ, random = ~1 | fBeach, method = "ML")
M1.REML <- lme(Richness ~ NAP + fExp, random = ~1 | fBeach, method = "REML", data = RIKZ)
```

<br><br>

The book provides a summary of the differences between the models:

![](`r REML2`){width=550}

<br><br><br><br><br><br><br><br><br><br>
















