---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nlme)
RIKZ <- read.table(here::here("data","RIKZ.txt"), stringsAsFactors = FALSE, header = TRUE)
RIKZ$fBeach <- factor(RIKZ$Beach)


DataStruc <- here::here("presentations/9-24-Presentation/images","DataStructure.jpg")
Stage21 <- here::here("presentations/9-24-Presentation/images","2stage1.jpg")
Stage22 <- here::here("presentations/9-24-Presentation/images","2stage2.jpg")
RandomInt <- here::here("presentations/9-24-Presentation/images","RandomIntercept.jpg")
RandomIntSlpe <- here::here("presentations/9-24-Presentation/images","RandomInterceptSlope.jpg")
```


<center>
## Mixed Effects Modelling {.tabset .tabset-fade}

![](`r DataStruc`){width=400}

<left>

### 2-Stage Analysis

```{r, echo=T}
Beta <- vector(length = 9)
for (i in 1:9)
{
  Mi <- summary(lm(Richness ~ NAP, subset = (Beach==i), data=RIKZ))
  Beta[i] <- Mi$coefficients[2,1]
}

fExposure9 <- factor(c(0, 0, 1, 1, 0, 1, 1, 0, 0))  # 1: Exposure == 11
                                                    # 0: Exposure != 11
Beta

tmp2 <- lm(Beta ~ fExposure9)

anova(tmp2)
```

Since the p-value for Exposure in the model for our slopes is 0.2268, we fail to find a significant exposure effect on the slopes.

<br>

![](`r Stage21`){width=400} <br>
![](`r Stage22`){width=400}

<br>

<font color="green"> Benefit <font color="black">
<br> 1. Very simple application of standard linear regression, yielding potentially useful results.

<br>

<font color="red"> Drawback <font color="black">
<br> 1. All of the data is being summarized in the second step by a single parameter.
<br> 2. We are analzing our regression parameters, meaning our understanding of the data is indirect.
<br> 3. There's no penalty (as is the case with Radj or AIC) for low sample sizes.

<br><br><br><br><br>

### Random Intercept Model

```{r, echo=T}
Mlme1 <- lme(Richness ~ NAP, random = ~1 | fBeach, data=RIKZ)

# Here, random = ~1 | fBeach specifies a random intercept model based on the nominal variable fBeach.

F0 <- fitted(Mlme1, level=0) # Fitted values for population
F1 <- fitted(Mlme1, level=1) # Fitted values for the within-beach subpopulations
I <- order(RIKZ$NAP); NAPs <- sort(RIKZ$NAP)

summary(Mlme1)

# In using beach as a "random effect," we can find the variance d for our intercepts in the "Random effects:" section of the summary statistics.

plot(NAPs, F0[I], lwd = 4, type = "l", ylim = c(0,22), ylab = "Richness", xlab = "NAP")

for (i in 1:9)
{
  x1 <- RIKZ$NAP[RIKZ$Beach == i]
  y1 <- F1[RIKZ$Beach == i]
  K <- order(x1)
  lines(sort(x1), y1[K])
}

summary(Mlme1)

```

<br> 

![](`r RandomInt`){width=400}

<br>

Since the variance for the random intercept is <font color="blue">2.94<font color="black">,<br>
and the variance for the residuals is <font color="blue">3.06<font color="black">,<br>
we can use these values to calculate the <b>induced correlation:</b>

2.94<sup>2</sup> / (2.94<sup>2</sup> + 3.06<sup>2</sup>) = 0.48

This is the correlation between observations from the same beach. We can use this value to calculate the design effect (1+(n-1)*p, where p = induced correlation), which helps us know whether 

<br><br><br><br><br>



### Random Intercept and Slope Model

```{r echo=FALSE, message=F, warning = F, warnings = F}
Mlme2 <- lme(Richness ~NAP, random = ~1 + NAP | fBeach, data = RIKZ)

F0 <- fitted(Mlme2, level=0)
F1 <- fitted(Mlme2, level=1)
I <- order(RIKZ$NAP); NAPs <- sort(RIKZ$NAP)
plot(NAPs, F0[I], lwd = 4, type = "l", ylim = c(0,22), ylab = "Richness", xlab = "NAP")

for (i in 1:9)
{
  x1 <- RIKZ$NAP[RIKZ$Beach == i]
  y1 <- F1[RIKZ$Beach == i]
  K <- order(x1)
  lines(sort(x1), y1[K])
}

summary(Mlme2)

```

<br>

![](`r RandomIntSlpe`){width=400}

<br>

### Random Effects Model

```{r, echo=FALSE}
Mlme3 <- lme(Richness ~ 1, random = ~1 | fBeach, data = RIKZ)
summary(Mlme3)
```

### Marginal Model

```{r, echo=FALSE}
M.mixed <- lme(Richness ~ NAP, random = ~1 | fBeach, metho = "REML", data=RIKZ)
M.gls <- gls(Richness ~ NAP, method = "REML", correlation = corCompSymm(form = ~1 | fBeach),
             data = RIKZ)

summary(M.gls$modelStruct)
```

### ML and REML

```{r, echo=FALSE}
RIKZ$fExp <- RIKZ$Exposure
RIKZ$fExp[RIKZ$Exposure == 8] <- 10
RIKZ$fExp <- factor(RIKZ$fExp, levels = c(10,11))
M0.ML <- lme(Richness ~ NAP, data = RIKZ, random = ~1 | fBeach, method = "ML")
M0.REML <- lme(Richness ~ NAP, random = ~1 | fBeach, method = "REML", data = RIKZ)
M1.ML <- lme(Richness ~ NAP + fExp, data = RIKZ, random = ~1 | fBeach, method = "ML")
M1.REML <- lme(Richness ~ NAP + fExp, random = ~1 | fBeach, method = "REML", data = RIKZ)

summary(M0.ML)
summary(M0.REML)
summary(M1.ML)
summary(M1.REML)
```



